---
title: "Package **_ci_**"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r README-0, include=FALSE}
options(bio.local_list = TRUE)

options(
  repos = "https://cran.rstudio.com/",
  pkgType = "both",
  install.packages.check.source = "yes",
  install.packages.compile.from.source = "always"
)

Sys.setenv(R_REMOTES_NO_ERRORS_FROM_WARNINGS = "true")
```


<!-- badges: start -->
[![GitHub version](https://img.shields.io/badge/GitHub-`r packageVersion("ci")`-brightgreen.svg)](https://github.com/GegznaV/ci)
[![R-CMD-check](https://github.com/GegznaV/ci/workflows/R-CMD-check/badge.svg)](https://github.com/GegznaV/ci/actions)
[![Updated-on](https://img.shields.io/badge/Updated%20on-`r gsub('-', '--', Sys.Date())`-yellowgreen.svg)](/commits/master)
<!-- badges: end -->



Convenience functions for calculating confidence intervals (CI).


Report bugs, issues, and desired improvements at https://github.com/GegznaV/ci/issues.
Please specify your operating system.


# Installation

Installation from a CRAN:
```{r README-1, eval=FALSE}
install.packages("ci")
```

## Development version

Installation from a CRAN-like repository:

```{r README-1a, eval=FALSE}
repos <- c("https://GegznaV.github.io/download/", getOption("repos"))
install.packages("ci", repos = repos)
```


<details>
<summary>Install from GitHub</summary>

Install from GitHub:

```{r README-2, eval=FALSE}
if (!require(remotes)) install.packages("remotes")
remotes::install_github("GegznaV/ci", dependencies = TRUE)
```

Note! To install from GitHub on Windows, you need the RTools.

</details>


# Examples

```{r README-5, paged.print=FALSE, eval=FALSE}
ci::ci_binom(x = 20, n = 101)
```
Or:
```{r README-6, paged.print=FALSE, eval=FALSE}
library(ci)
ci_binom(x = 20, n = 101)
```


```{r README-7, paged.print=FALSE, eval=FALSE}
library(tidyverse)
library(ci)

data(npk, package = "datasets")
head(npk)

npk |>
  group_by(N, P, K) |> 
  ci_mean_t(yield)
```
